---
title: "REM Sleep Analysis"
author: "Devon Mastin"
date: "`r Sys.Date()`"
output: 
  html_document:
    self-contained: TRUE
    df_print: paged
    toc: TRUE
    toc.position: left
editor_options:
  markdown:
    wrap: sentence
---

## Introduction

This page will run some simple analyses a data set included in ggplot2 called msleep. 
```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(leaflet)
library(hexbin)
library(tinytex)
library(multcomp)
library(multcompView)
library(ggcorrplot)
library(viridis)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## Long Data Frame
```{r}
animalsleep<- dplyr::select(msleep, "name", "sleep_total")
kable(animalsleep, col.names = c("Common Name", "Total Sleep (h)"))
```

This is a long data frame with many many rows. Nobody wants to look at this. Lets see if we can get a better look. 



## Visualization
```{r sleep ggplot, fig.align='center'}
msleep%>%
  filter(!is.na(vore))%>%
  ggplot(aes(x = sleep_total, y = sleep_rem, colour = vore))+
  geom_point()+
  xlab("Total Sleep (h)")+
  ylab("REM Sleep (h)") +
  geom_smooth(method = 'lm', se = FALSE)+
  labs(
    title = "Total Sleep vs REM Sleep",
    caption = "any rows with a na value for vore have been filtered out"
  )
```
Cool! 
Lets look at another relationship...

```{r 2ggplot, fig.align='center'}
msleep%>%
  ggplot()+
  geom_point(mapping= aes(x=bodywt, y=sleep_total)) +
  labs(
    title = "Body Weight (g) vs Total Sleep (h)",
    x = "Body Weight (g)",
    y = "Total Sleep (h)"
  )
```

It doesn't look like there is much of a relationship here but let's filter out those larger values to get a better picture.




```{r 2ggplot(2), fig.align='center'}
msleep%>%
  filter(bodywt < 2000)%>%
  ggplot()+
  geom_point(mapping=aes(x=bodywt, y=sleep_total)) +
  labs(
    title = "Body Weight (g) vs Total Sleep (h)",
    x = "Body Weight (g)",
    y = "Total Sleep (h)"
  )
```

Maybe, but not really clear. Lets run a statistical analysis on the data (bringing back in those values that we filtered out).



```{r 1correlation coefficient}
correlation <- cor(msleep$bodywt, msleep$sleep_total, method='pearson')
print(paste("pearson correlation coefficient:", correlation))
```

Ok, so our correlation between those two variables is pretty weak, which makes sense when we look at our graph. Let's look for a strong correlation somewhere else. 





## Correlation Matrix



I am using a cool package called ggcorrplot which lets me make a visual correlation matrix, though when I tried this with all the data it was a huge mess. So I selected all the numeric data and ran the matrix.



```{r correlation matrix, fig.align='center'}
sleepdata <- msleep%>%
  select_if(is.numeric)

model.matrix(~0+., data=sleepdata)%>%
  cor(use="pairwise.complete.obs")%>%
  ggcorrplot(show.diag=FALSE, type="lower", lab=TRUE, lab_size =0.005)
```

It shows an interesting relationship I wouldn't have thought to check for between sleep cycle time and brain weight.


Lets look a little closer at this. 





## Brain Weight and Sleep Cycle Time
```{r new relationship, fig.align='center'}
ggplot(msleep, aes(x=brainwt, y=sleep_cycle, colour=vore))+
  geom_point()+
  labs(
    title = "Brain Weight vs Sleep Cycle Time (h)",
    x = "Brain Weight (g)",
    y = "Sleep Cycle Time (h)"
  )
```





```{r}
cleansleep <- msleep%>%
  filter(!is.na(brainwt))%>%
  filter(!is.na(sleep_cycle))

correlation_coefficient <- cor(cleansleep$brainwt, cleansleep$sleep_cycle)
p_value <- cor.test(cleansleep$brainwt, cleansleep$sleep_cycle)$p.value
print(paste("correlation_coefficient:", correlation_coefficient))
print(paste("P-value:", p_value))
```





## Order and Sleep

Let's look at whether the order affects the total sleep time and REM sleep. There are a lot of orders so I am just going to look at Rodentia and Carnivora.

```{r new data set, fig.align='center'}
sleep<-filter(msleep, order=='Rodentia'|order=='Carnivora')%>%
  filter(sleep_rem!="")
```
```{r}
ggplot(sleep, aes(x=sleep_total, y=sleep_rem, colour=order))+
  geom_point()+
  geom_smooth(method='lm', se=FALSE)+
  theme_bw()
```





Let's run a linear model to ask that question from above; is the effect of total sleep on REM sleep different for the orders Rodentia and Carnivora?

```{r sleep lm, echo = TRUE}
model1<-lm(sleep_rem~sleep_total*order, data=sleep)
summary(model1)
```

We looked at whether total sleep, order, or the interaction between the two variables had a determining factor on how much REM sleep the animal got. When we look at our p-values it looks like total sleep did have a significant effect (shocker!), but order, and the combined effect of order and total sleep were not significant factors. 




Let's make a new data set where REM sleep is a function of total sleep or order, but total sleep and order are not dependent on eachother, then model it.

```{r, echo = TRUE}
model2 <- lm(sleep_rem~sleep_total+order, data=sleep)
summary(model2)
```

This summary shows us that total sleep does create a significant difference in the amount of REM sleep, but that order of the animal was not a significant predictor of REM sleep.




Now let's look at whether the two models are significantly different from each other.

```{r}
anova(model1, model2)
```


The two models are not significantly different from each other as evidenced by our high p-value, and with our first model we lose a degree of freedom, so the second model may be a better option.



 
We want to add in a column that shows a predicted lm value based on the second model then visualize it.


```{r, echo = TRUE, fig.align='center'}
sleep$predlm<-predict(model2)
ggplot(sleep, aes(x=sleep_total, y=sleep_rem, colour=order))+
  geom_point()+
  geom_line(aes(y=predlm))
```



OK that was fun, go watch this video below before you go


## Cute Video

<iframe width="610"
height="348"
src="https://youtu.be/EPrTHSzUPFE?si=gTjwdEAJPcSuOn6u"
date-external="1">

</iframe>
